\documentclass{article}

% NeurIPS 2026 Position Paper Track submission format.
% Do not add "final" or "preprint" for an anonymized submission.
\usepackage[position]{neurips_2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}

\newcommand{\method}{\textsc{adapt}}

\title{Static Sparsity Is the Wrong Unit of Evaluation for\\
Inference-Time Attention Pruning}

\author{%
  Anonymous Author(s) \\
  Affiliation \\
  Address \\
  \texttt{email} \\
}

\begin{document}

\maketitle

\begin{abstract}
Structured attention pruning is usually evaluated by aggregate sparsity and
average task accuracy. This framing treats a pruned transformer as if its
computational budget were a fixed property of the model rather than a decision
made for each input. We argue that this is the wrong abstraction for reasoning
workloads. A static pruning mask and an input-adaptive pruning policy can have
the same average keep ratio while allocating compute in very different ways.
The evaluation question should therefore shift from ``which heads can be
removed globally?'' to ``does the system allocate attention heads appropriately
for the input currently being processed?'' We support this position with a
case study of activity-dependent attention pruning, \method{}, in which small
learned agents produce per-input attention-head masks from runtime activation
statistics. Across two full BIG-Bench Hard sweeps with Qwen3-4B at matched
reported keep ratio, \method{} improves over a static mask by 9.9 and 9.5
percentage points in mean task accuracy. These results do not establish a final
benchmark claim, but they show that static and adaptive pruning are not
interchangeable under aggregate sparsity. We propose that future work report
per-input keep-ratio distributions, adaptive-static matched-sparsity gaps, and
mask variability diagnostics alongside standard accuracy and latency.
\end{abstract}

\section{Introduction}

Transformer inference uses the same attention structure for every input. A
short factual query, a simple arithmetic problem, and a long compositional
reasoning prompt all activate every attention head in every layer. The standard
response to this inefficiency is structured pruning: estimate which attention
heads are globally redundant, remove or mask them, and evaluate the resulting
model by its average sparsity and task accuracy
\citep{michel2019sixteen,voita2019analyzing,clark2019does}.

That evaluation protocol is convenient, but it hides the most important
decision made by an inference system: how much compute to spend on a particular
input. A fixed pruning mask commits to the same subnetwork for all examples.
An adaptive pruning policy can keep more capacity for hard inputs and suppress
more capacity for easy ones. These two systems may have the same average
sparsity, but they implement different computational policies.

This paper argues that aggregate sparsity is an insufficient unit of evaluation
for inference-time pruning. The central object of study should be per-input
compute allocation. In particular, pruning papers should not only report the
mean fraction of heads kept; they should report how that fraction varies across
examples, whether adaptivity changes behavior under matched sparsity, and
which inputs benefit or suffer from dynamic allocation.

Our position is motivated by reasoning workloads. In tasks such as symbolic
deduction, mathematical word problems, and multi-step question answering, the
computational demand of an input is not well captured by dataset identity alone.
Even within a single benchmark, examples vary in the number of latent steps,
the amount of context that must be maintained, and the degree to which
intermediate facts must be combined. A globally fixed attention mask is a poor
fit for that variation.

We make three claims.

\begin{enumerate}
  \item Static attention pruning encodes a stronger assumption than is usually
        acknowledged: it assumes head utility is stable enough across inputs
        that one global mask is an adequate inference policy.
  \item Matched average sparsity does not imply matched computation. Adaptive
        and static masks can have the same mean keep ratio while preserving
        different heads on different examples.
  \item Evaluation protocols for efficient reasoning models should include
        per-input allocation diagnostics, not only average sparsity, accuracy,
        and latency.
\end{enumerate}

We use \method{}, an activity-dependent pruning system, as a case study rather
than as the sole contribution of the paper. \method{} pairs each transformer
layer with a small learned agent that maps runtime activation statistics to
head-level keep masks. The system is inspired by a simple biological principle:
microglia refine neural circuits in response to local patterns of activity
\citep{stevens2007complement,paolicelli2011synaptic}. We do not claim a
one-to-one biological model. The useful abstraction is activity-dependent
allocation: connections are evaluated in context, not only by a global
importance score.

\section{Position}

The prevailing question in attention pruning is: which heads are dispensable?
This question is underspecified. A head may be dispensable for most examples
and still be useful for a minority of hard cases. Conversely, a head may be
useful on average while contributing little to a large set of easy inputs. A
single global answer collapses this variation.

We propose a different question: which heads are necessary for this input? This
question treats pruning as an inference-time allocation problem. It does not
require that all pruning methods become dynamic, but it does require that
dynamic behavior be measured directly when it is claimed. A model that keeps
75\% of heads on average could keep exactly the same 75\% for every input, or
it could keep 95\% for hard inputs and 55\% for easy ones. Reporting only the
mean keep ratio makes these systems appear equivalent.

This matters because efficiency is not a scalar property of a model alone. It
is a property of a model, an input distribution, and an allocation policy. For
reasoning models, the input distribution often contains a mixture of trivial,
moderate, and difficult cases. A useful efficient system should exploit that
mixture.

\section{Why Static Masks Are Incomplete}

Static pruning has clear advantages. It is simple, easy to deploy, and directly
compatible with optimized kernels when heads are physically removed. It also
has an attractive experimental protocol: compute an importance score for each
head, prune the lowest-scoring heads, and measure the resulting accuracy and
latency. This is a reasonable baseline.

The limitation is that static pruning turns a distributional statistic into an
input-level decision. Let $m_{l,h} \in \{0,1\}$ denote a static mask for head
$h$ in layer $l$. Once chosen, $m_{l,h}$ is applied to every example. The mask
therefore represents an average judgment about utility:
\[
  m_{l,h} = f\left(\mathbb{E}_{x \sim \mathcal{D}}[u_{l,h}(x)]\right),
\]
where $u_{l,h}(x)$ is an unobserved, input-dependent utility. If the variance
of $u_{l,h}(x)$ is small, this approximation is acceptable. If the variance is
large, a static mask can prune heads that are rarely but critically useful.

Dynamic inference methods already recognize this issue in other forms.
Conditional computation allocates model capacity as a function of the input
\citep{bengio2015conditional}.
Early-exit models allocate fewer layers to easy examples
\citep{graves2016adaptive,schuster2022confident}. Token-pruning methods reduce
sequence computation based on token salience
\citep{goyal2020power,wang2021spatten}. Mixture-of-experts models route inputs
through different feed-forward experts \citep{shazeer2017outrageously,fedus2022switch}.
Head-level pruning should be evaluated with the same lens: the policy matters,
not only the average number of active units.

\section{An Evaluation Protocol for Adaptive Pruning}

We recommend that pruning papers report four quantities when evaluating
input-adaptive methods.

\paragraph{Matched-sparsity adaptive-static gap.}
Compare an adaptive policy against a static policy at the same average keep
ratio. This is the primary test of whether adaptivity adds value beyond simply
using more compute.

\paragraph{Per-input keep-ratio distribution.}
Report the distribution of keep ratios across examples, not just the mean. If a
method claims to allocate compute dynamically, the distribution should show how
budgets vary.

\paragraph{Stratified performance.}
Evaluate easy and hard subsets separately. A useful adaptive policy should not
only improve average accuracy; it should reveal where additional compute is
being spent and where it is being saved.

\paragraph{Mask variability diagnostics.}
Report whether masks vary across examples within a task and across tasks. Mean
mask similarity alone is not sufficient. A high similarity between task-average
masks can coexist with meaningful per-example variation, and a low similarity
does not by itself prove improved allocation.

These diagnostics should be treated as part of the empirical claim, not as
optional interpretability figures. Without them, aggregate sparsity can obscure
the behavior that adaptive pruning is meant to provide.

\section{Case Study: Activity-Dependent Attention Pruning}

To make the position concrete, we study \method{}, a simple input-adaptive
attention-pruning framework. The goal is not to introduce a complex new
architecture. The goal is to instantiate the evaluation problem in a way that
allows static and adaptive policies to be compared under matched sparsity.

Consider a transformer with $L$ layers and $H$ attention heads per layer. For
an input $x$, \method{} predicts a soft keep mask
$m_{l,h}(x) \in [0,1]$ for each head. The attention output at layer $l$ becomes
\[
  \hat{A}_l(x) =
  \operatorname{Concat}\left[m_{l,1}(x)a_{l,1}, \ldots,
  m_{l,H}(x)a_{l,H}\right] W_l^O,
\]
where $a_{l,h}$ is the output of head $h$ and $W_l^O$ is the output
projection.

Each layer has a small two-layer agent network. The agent receives activation
statistics computed during the forward pass, including head activation
magnitude and attention entropy. These features are inexpensive summaries of
whether a head is sharply engaged, diffusely attending, or weakly active. The
agent maps the layer-level feature vector to $H$ mask logits, which are passed
through a temperature-scaled sigmoid.

The training objective balances task fidelity and sparsity:
\[
  \mathcal{L} =
  \mathcal{L}_{\text{task}} +
  \alpha \mathcal{L}_{\text{sparse}} +
  \beta \mathcal{L}_{\text{binary}}.
\]
The task loss preserves answer accuracy. The sparsity loss penalizes average
mask value. The binary regularizer discourages indecisive masks near 0.5.
During training, sparsity pressure is increased gradually so that agents first
learn which heads support the task and only later learn to suppress weak heads.

This system is intentionally modest. It is a probe of an evaluation thesis:
when average keep ratio is held fixed, does an input-conditioned mask behave
differently from a static mask?

\section{Preliminary Evidence on BIG-Bench Hard}

We ran two full BIG-Bench Hard sweeps \citep{suzgun2022challenging} using
Qwen3-4B-Instruct-2507. Each sweep covered 27 BBH tasks with 25 examples per
task. We compared three conditions: the unpruned model, a static head mask, and
\method{}. Static and adaptive conditions were evaluated at a reported mean keep
ratio of 0.75. The two sweeps used different random seeds.

\begin{table}[t]
  \centering
  \caption{Mean task accuracy across 27 BIG-Bench Hard tasks. Each task uses
  25 examples. The adaptive-static gap is computed after averaging task
  accuracies within each run.}
  \label{tab:bbh}
  \begin{tabular}{lcccccc}
    \toprule
    Run & Seed & Unpruned & Static & \method{} & \method{} $-$ Static &
    Tasks improved \\
    \midrule
    BBH sweep 1 & 42 & 0.339 & 0.201 & 0.301 & +0.099 & 17 / 27 \\
    BBH sweep 2 & 43 & 0.339 & 0.098 & 0.193 & +0.095 & 14 / 27 \\
    \bottomrule
  \end{tabular}
\end{table}

The result is not that \method{} closes the gap to the unpruned model. It does
not. The result is that the adaptive policy consistently improves over the
static policy under the same reported keep ratio. Across the two sweeps,
\method{} improves mean task accuracy by 9.9 and 9.5 percentage points over
static pruning.

The task-level pattern is also informative. In one sweep, the largest gains
appear on \texttt{salient\_translation\_error\_detection},
\texttt{hyperbaton}, \texttt{causal\_judgement}, \texttt{navigate}, and
\texttt{disambiguation\_qa}. These are not proof of a taxonomy, since each
task has only 25 examples, but they illustrate why a single average can be
misleading. Some tasks show large adaptive-static gaps; others show no
difference; one task in the second sweep favors static pruning.

The mask-similarity analysis is more cautionary. Mean adaptive masks across
BBH tasks are highly similar under cosine similarity. This does not support a
claim that \method{} learns a sharply different average mask for each task.
Instead, the useful conclusion is narrower: adaptive pruning can outperform a
static mask even when task-average masks appear similar. Future analysis should
therefore focus on per-example variation, not only task-level average masks.

These results are preliminary. They use two seeds and a small number of
examples per task. They are nevertheless sufficient for the position advanced
here: adaptive and static pruning should not be treated as equivalent merely
because their aggregate keep ratios match.

\section{What Would Falsify This Position?}

A useful position paper should state what evidence would weaken its claim. Our
argument would be undermined by any of the following findings.

\begin{itemize}
  \item Across many tasks, seeds, and model families, adaptive masks provide no
        improvement over strong static masks at matched keep ratio.
  \item Per-input keep ratios collapse to a constant, indicating that the
        adaptive mechanism is dynamic in form but static in behavior.
  \item Adaptive gains disappear when static masks are calibrated with stronger
        importance estimates or larger calibration sets.
  \item Reported adaptive gains are explained entirely by implementation
        artifacts, such as mismatched sparsity accounting or different decoding
        settings.
\end{itemize}

These are not minor concerns. They define the next experiments that should be
run. A position in favor of per-input allocation does not exempt adaptive
methods from strong static baselines.

\section{Implications}

If per-input allocation is the right object of study, several practices should
change.

First, matched-sparsity baselines should be standard. Adaptive methods should
not be compared only to the full model or to weak random pruning. The right
baseline is a carefully calibrated static mask with the same average keep
ratio.

Second, papers should report allocation behavior. Mean sparsity is not enough.
Histograms of keep ratio, mask variance within tasks, and easy-hard
stratification should be included whenever dynamic pruning is claimed.

Third, latency claims should separate algorithmic sparsity from realized
speedup. A dynamic mask that improves accuracy at a fixed compute budget is
scientifically interesting even before it is fully optimized for hardware.
Conversely, a method that reports sparsity without kernel-level speedup should
be clear about that distinction.

Fourth, adaptive pruning should be evaluated as a policy, not only as a
compressed model. The policy decides where compute is spent. That decision is
part of the method.

\section{Limitations}

The empirical evidence here is intentionally modest. The BBH sweeps use 25
examples per task and two seeds. This is enough to motivate the evaluation
claim, but it is not enough to establish a definitive benchmark result.

The current case study also uses one model family. Larger models may exhibit
different redundancy patterns, and stronger static pruning methods may reduce
the observed adaptive-static gap. The static baseline must therefore be
strengthened before making broad performance claims.

The biological analogy should also be kept in perspective. Microglial pruning
motivates the activity-dependent framing, but transformers do not implement
synaptic pruning, complement signaling, or biological plasticity. The analogy
is useful as a design principle, not as a mechanistic claim.

Finally, adaptive masks create deployment challenges. Static pruning can be
compiled into a smaller model. Dynamic pruning requires runtime decisions and
hardware support to turn those decisions into latency gains. For this reason,
accuracy at matched sparsity and realized speedup should be reported
separately.

\section{Conclusion}

Attention pruning should not be evaluated only as a search for globally
dispensable heads. For reasoning workloads, the more relevant question is how
compute is allocated across inputs. A static mask and an adaptive policy can
share the same average keep ratio while making different decisions on the
examples that matter.

Our case study provides preliminary evidence for this distinction. Across two
full BIG-Bench Hard sweeps, an activity-dependent pruning policy improves over
a static mask by roughly ten percentage points in mean task accuracy at matched
reported keep ratio. The result is not a final claim of state-of-the-art
efficiency. It is evidence that aggregate sparsity is too coarse to evaluate
adaptive inference.

Future work on pruning should report per-input allocation behavior alongside
accuracy, sparsity, and latency. Without those diagnostics, the field risks
treating very different inference policies as equivalent because their averages
look the same.

\section*{References}

{\small
\begin{thebibliography}{99}

\bibitem[Bengio et al.(2015)]{bengio2015conditional}
Bengio, E., Bacon, P.-L., Pineau, J., and Precup, D. (2015).
\newblock Conditional computation in neural networks for faster models.
\newblock \emph{arXiv preprint arXiv:1511.06297}.

\bibitem[Clark et al.(2019)]{clark2019does}
Clark, K., Khandelwal, U., Levy, O., and Manning, C. D. (2019).
\newblock What does BERT look at? An analysis of BERT's attention.
\newblock In \emph{Proceedings of the 2019 ACL Workshop BlackboxNLP}.

\bibitem[Fedus et al.(2022)]{fedus2022switch}
Fedus, W., Zoph, B., and Shazeer, N. (2022).
\newblock Switch transformers: Scaling to trillion parameter models with
simple and efficient sparsity.
\newblock \emph{Journal of Machine Learning Research}, 23(120):1--39.

\bibitem[Goyal et al.(2020)]{goyal2020power}
Goyal, S., Choudhury, A., Raje, S., Chakaravarthy, V., Sabharwal, Y., and
Verma, A. (2020).
\newblock PoWER-BERT: Accelerating BERT inference for classification tasks.
\newblock In \emph{International Conference on Machine Learning}.

\bibitem[Graves(2016)]{graves2016adaptive}
Graves, A. (2016).
\newblock Adaptive computation time for recurrent neural networks.
\newblock \emph{arXiv preprint arXiv:1603.08983}.

\bibitem[Michel et al.(2019)]{michel2019sixteen}
Michel, P., Levy, O., and Neubig, G. (2019).
\newblock Are sixteen heads really better than one?
\newblock In \emph{Advances in Neural Information Processing Systems}.

\bibitem[Paolicelli et al.(2011)]{paolicelli2011synaptic}
Paolicelli, R. C. et al. (2011).
\newblock Synaptic pruning by microglia is necessary for normal brain
development.
\newblock \emph{Science}, 333(6048):1456--1458.

\bibitem[Schuster et al.(2022)]{schuster2022confident}
Schuster, T., Fisch, A., Jaakkola, T., and Barzilay, R. (2022).
\newblock Confident adaptive language modeling.
\newblock In \emph{Advances in Neural Information Processing Systems}.

\bibitem[Shazeer et al.(2017)]{shazeer2017outrageously}
Shazeer, N. et al. (2017).
\newblock Outrageously large neural networks: The sparsely-gated
mixture-of-experts layer.
\newblock In \emph{International Conference on Learning Representations}.

\bibitem[Stevens et al.(2007)]{stevens2007complement}
Stevens, B. et al. (2007).
\newblock The classical complement cascade mediates CNS synapse elimination.
\newblock \emph{Cell}, 131(6):1164--1178.

\bibitem[Suzgun et al.(2022)]{suzgun2022challenging}
Suzgun, M. et al. (2022).
\newblock Challenging BIG-bench tasks and whether chain-of-thought can solve
them.
\newblock \emph{arXiv preprint arXiv:2210.09261}.

\bibitem[Voita et al.(2019)]{voita2019analyzing}
Voita, E., Talbot, D., Moiseev, F., Sennrich, R., and Titov, I. (2019).
\newblock Analyzing multi-head self-attention: Specialized heads do the heavy
lifting, the rest can be pruned.
\newblock In \emph{Proceedings of ACL}.

\bibitem[Wang et al.(2021)]{wang2021spatten}
Wang, H., Zhang, Z., and Han, S. (2021).
\newblock SpAtten: Efficient sparse attention architecture with cascade token
and head pruning.
\newblock In \emph{IEEE International Symposium on High-Performance Computer
Architecture}.

\end{thebibliography}
}

\newpage
\input{checklist.tex}

\end{document}
